---
title: "Census Data Wrangling and Mapping in R - Tutorial"
author: "Patty Frontiera"
date: "11/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# R Census Data - Tutorial Time!

## Part 1

We will work through several exercises using `tidycensus` to fetch, wrangle, and map census data.

## Setup

Be sure to **clone or downloaded and unzip** the workshop files from: [https://github.com/dlab-berkeley/Census-Data-in-R](https://github.com/dlab-berkeley/Census-Data-in-R)

Then set your working directory to the folder that contains the unzipped files.

You can do this by:

1. Open the folder with the workshop files

2. Double-click on the R Project file `Census-Data-in-R.Rproj`

3. This should open RStudio - with the `Files` panel displaying the workshop folder contents.

4. Double-click on the file `Rcensus_data_maps-tutorial.Rmd` to follow along!

*You can also click on the file `Rcensus_data_maps-tutorial.html` in the Files tab to open the workshop tutorial in a web brower.*

## Install packages

Next, install any packages we will use that are not already installed on your system.

- If you installed any of these awhile ago it's a good idea to install updates (though not during the workshop as things can break!)

Basic syntax to install an R package:

> `install.packages("package_name")`

Here is some code to install a set of R packages.

```{r}
# Create a list of the packages we will use
list_of_packages <- c("tidyverse","sf","mapview","tigris", "tidycensus")

# Identify any new packages that will need to be installed
new_packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]

# Install any packages that are not installed (new_packages)
if(length(new_packages) > 0) {
  print(paste("Installing these packages:", new_packages))
  install.packages(new_packages)
} else {
  print("All packages already installed!")
}
```


## Loading packages

Use the `library` function to load the R packages we will use today

```{r, message=FALSE, warning=FALSE}
library(tidycensus)
library(tidyverse) 
#library(tigris)
library(sf)
library(mapview)
```

*If you are getting errors try loading the `dplyr` package or reinstalling it as that has worked for some.*

## Census API Key

You need a `Census API key` to programmatically fetch census data.

- Get it here (pretty quickly): <https://api.census.gov/data/key_signup.html>

- The key will be sent to your email and you will need to click to activate it.

- Keep the email with the key open for use in this workshop.

*For more info on all available Census APIs see: <https://www.census.gov/data/developers/data-sets.html>*

## Census Data with the `tidycensus` R package

The [tidycensus](https://walker-data.com/tidycensus/index.html) package allows R users to quickly fetch data from a select subset of Census databases.

- Many of the code samples in this tutorial borrow heavily from [Kyle Walker's](https://walker-data.com) tidycensus package documentation.

The key tidycensus functions we will use today are:

- `census_api_key`: makes your Census API key available to tidycensus

- `load_variables`: retrieves a dataframe of available census data variables

- `get_decennial`: fetch census data for the most recent decennial censuses - 2000, 2010 (and soon 2020)

- `get_acs`: fetch 2005 - 2020 data from the 1 and 5 year ACS (American Community Survey) data bases


## Add Your Census API Key

Let's get started! 

1. Copy and paste your Census API key from your email

1. Use the tidycensus function **census_api_key** to register your API key with tidycensus - **don't forget to put quotes around the key!**.

```{r, eval=F}
# Install your census api key - long alphanumeric string
census_api_key("THE_BIG_LONG_ALPHANUMERIC_API_KEY_YOU_GOT_FROM_CENSUS")
```

## Another way to add your Census API Key

I keep my key in a file so no one can see it
```{r}

# source (run) an r script that creates a variable with my key
source("../../keys/census_api_key.R")

#print(my_census_api_key) 

# register the key
census_api_key(key = my_census_api_key)
```


# Fetching Decennial Census Data

## `get_decennial` function

Let's start by fetching from the 2010 Census with tidycensus's `get_decennial` function.

```{r}
pop2010 <- get_decennial(geography = "state",   # census tabulation unit
                         variables = "P001001", # variable(s) of interest
                         year = 2010)           # census year

head(pop2010)          
```


## tidycensus data is `tidy`

By default, tidycensus returns data in a `tidy`, or `long` format that allows data for multiple variables to be contained within the `variable` and `value` columns. This is in contract to untidy, or `wide` data where each variable is in its own column. 

tidycensus can return long data if you can add the parameter `output=wide` to the function call.
  
```{r}
# wide format
pop2010w <- get_decennial(geography = "state",   # census tabulation unit
                         variables = "P001001", # variable(s) of interest
                         year = 2010,           # census year
                         output="wide")         # get output in wide format
head(pop2010w) 
```

## The `GEOID` column

The `GEOID` column is included in tidycensus output. 

This is a Census geographic identifier for the tabulation unit.

The `GEOID` is sometimes called the Census `FIPS` code and for most tabulation units these are the same. 

- This makes it possible to link to Census geographic data and make maps. (We will do this in a bit.)

The `GEOID` is a text string and must be quoted. 

- Beware of GEOID leading zeros, since some software will remove these and convert GEIOD values to numbers (rather than text strings).

#### Question:

*What is the GEOID for California*

## Census Tabulation Units

Decennial census data is gathered from individuals and publicly distributed in aggregated form to protect privacy.

The Census tabulation units are the Census geographies to which the census data have been aggregated.

Some of the most common geographic tabulation units and their tidycensus function abbreviations are shown below, along with **required** and available filters that limite what data are returned.

|Geography    |Definition          |Filter(s)    |Used in tidycensus        |
|-------------|--------------------|-------------|--------------------------|
|"us"         |United States       |             |get_acs(), get_decennial()|
|"region"     |Census region       |             |get_acs(), get_decennial()|
|"state"      |State or equivalent |state        |get_acs(), get_decennial()|
|"county"     |County or equivalent|state, county|get_acs(), get_decennial()|
|"place"      |Census place        |state        |get_acs(), get_decennial()|
|"tract"      |Census tract        |**state**, county|get_acs(), get_decennial()|
|"block group"|Census block group  |**state**, county|get_acs(), get_decennial()|
|"block"      |Census block        |**state**, **county**|get_decennial() only! |

## `get_decennial` Geographic Tabulation Units and Filters

Let's work together to fill in the code to fetch state population in 2010 just for California. You can find the code in the file `Solutions.R`.

- See `?get_decennial` for help

```{r, eval=F}
#Fill in the code to fetch State population in 2010 just for California
get_decennial(geography = "state",               # census tabulation unit
                          variables = "P001001", # variable(s) of interest
                          year = 2010,           # census year
                          ...........)           # Filter by state is CA
```

Alter the above code to fetch state pop in 2010 for CA, TX and FL
```{r}
# Your code below or see Solutions.R
## fetch state pop in 2010 for CA, TX and FL

```

## Change the tabulation unit

Let's fetch 2010 population data for CA counties 

- *What changes in the code?*

```{r}
get_decennial(geography = "county",              # census tabulation unit
                          variables = "P001001", # variable(s) of interest
                          year = 2010,           # census year
                          state='CA')            # Filter by state is CA
```
#### Questions

- How do we specify the state of CA above? How else can we?

- Can you fetch population data for all counties in the USA or do you need to have a `state=` filter?

## Adding a county filter

You can also filter tidycensus results by `county`
```{r}
get_decennial(geography = "county",              # census tabulation unit
                          variables = "P001001", # variable(s) of interest
                          year = 2010,           # census year
                          state='CA',            # Filter by state is CA
                          county='Alameda')      # Filter by county Alameda
```

Alter the code above to fetch 2010 population for Alameda & San Francisco Counties
```{r}
# Your code below or see Solutions.R
## fetch 2010 population for Alameda & San Francisco Counties

```

## Visualize Results

We can visualize the data to get a quick overview of the distribution of data values.

It's a first step in exploratory data analysis and a last step in data communication.

`ggplot2` is the most commonly used R package for data visualization. 

- It is loaded when you load the `tidyverse` package.

Let's use it to visualize the population data.

## Plot 2010 Population by state

Use `ggplot2` to create an ordered horizontal bar chart.
```{r}
pop_plot<- ggplot(data=pop2010, aes(x=reorder(NAME,value), y=value/1000000)) + 
  geom_bar(stat="identity") + coord_flip() +
  theme_minimal() + 
  labs(title = "2010 US Population by State") +
  xlab("State") +
  ylab("in millions")

# display the plot
pop_plot
```

## Identifying Census Variables

In the code above we fetched data for total population in 2010 using the variable `"P001001"`.

That is not an obvious variable name, so how do we get those identifiers?

We can use the tidycensus **load_variables** function for this.


## load_variables

Use `load_variables` to fetch all variables used in the 2010 census into a dataframe.
```{r}
vars2010 <- load_variables(year=2010,        # Year or end year for ACS-5yr
                           dataset = 'sf1',  # 'sf1' for decennial or 'acs5', etc
                           cache = TRUE)     # Whether to save fetched data locally

# How large is the output
dim(vars2010)

# Take a look with head or View
head(vars2010)

```

## 2010 Decennial Census Tables & Variables

- Over 3,000 unique variables that describe population and housing characteristics

- Organized in `333` Tables
    - 177 population tables (identified with a ‘‘P’’) available to the block level 
    - 58 housing tables (identified with an ‘‘H’’) available to the block level
    - 82 population tables (identified with a ‘‘PCT’’) available to the census tract level
    - 4 housing tables (identified with an “HCT”) available to the census tract level
    - 10 population tables (identified with a “PCO”) available to the county level 
    - plus 2 additional PCT tables

<small>https://www.census.gov/data/datasets/2010/dec/summary-file-1.html</small>

## What Variable Has the 2010 Total Population value? 

We can sort and filter the vars2010 dataframe to find it.

### Questions

What 2010 decennial census variable contains...

- `Median Age` 

- `Average Family Size`

- `Number of occupied housing units` 

*Answers are in Solutions.R*

## Challenge

Complete the code cell below (...) to fetch and plot `Avg Family Size` by CA County in `2010`.

```{r, eval=F}
# Fetch Avg family size by CA County
ca_fam_size <- get_decennial(...)

# plot it
ca_fam_size %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()
```

## Challenge

Repeat the previous challenge with data from the `2000` decennial census.

Don't assume variable names are the same across the 2000 and 2010 census

- Use `load_variables` to check!

*See Solutions.R for the code*

```{r, eval = F}
# Add your code below


```

## Census Tract Data

Census tracts are the most commonly used census tabulation unit.

Let's fetch population data for the census tabulation unit to *tract*

> Because of the large number of census tracts, you MUST specify a state when requesting these data with tidycensus.


## Census Tract Data

Fetch total population for California by census tract

```{r, eval=F}
## Fetch population by **tract** for California.
ca_tract_pop2010 <- get_decennial(geography = "tract",   # census tabulation unit
                                   variables = "P001001", # variables of interest
                                   year = 2010,           # census year
                                   state='CA')      # State filter

# How many tracts in CA
dim(ca_tract_pop2010)

# take a look
head(ca_tract_pop2010)
```

## Fetching Census Tract Data

If you want census data at the tract level or below you **must** specifiy the state(s)

- You can also specify one or more counties
```{r,}
tract_pop2010 <- get_decennial(geography = "tract",   # census tabulation unit
                         variables = "P001001",       # variable of interest
                         year = 2010,                 # census year - only one!
                         state="CA",                  # limit to California
                         county=c("Alameda","Contra Costa"))  # & these counties
```




## Fetching data for more than one census variable

What **three** things are new here?
```{r}
#urban and rural pop for 3 CA counties
ur_pop10 <- get_decennial(geography = "county",  # census tabulation unit
                           variables = c(urban="P002002",rural="P002005"),
                           year = 2010, 
                           summary_var = "P002001",  # The denominator
                           state='CA',
                           county=c("Napa","Sonoma","Mendocino"))

```

## Fetching data for more than one census variable

1. You can specify more than one variable:
```
variables = c("P002002","P002005")
```

2. You can rename the values in the output 'variable' column.
```
variables = c(urban="P002002",rural="P002005")
```

3. You can identify a `summary_var` (a denominator - here, the total count of all people or households surveyed. Can be used for calculations like percent of total.) 
```
summary_var = "P002001"
```

## Take a look at the results
```{r}
ur_pop10
```

## Calculating Percents

The `summary_value` column comes in handy when you want to compute percent of total, for example:

```{r}
# Calculate the percent of population that is Urban or Rural
ur_pop10 <- ur_pop10 %>%
            mutate(pct = 100 * (value / summary_value))

```

## Calculating Percents

Let's take a look at the output.
```{r}
ur_pop10 # Take a look
```

## Plot it

Plots give us compact visual summaries of the data.
```{r}
myplot <- ggplot(data = ur_pop10, 
          mapping = aes(x = NAME, fill = variable, 
                     y = ifelse(test = variable == "urban", 
                                yes = -pct, no = pct))) +
          geom_bar(stat = "identity") +
          scale_y_continuous(labels = abs, limits=c(-100,100)) +
          labs(title="Urban & Rural Population in Wine Country", 
               x="County", y = " Percent of Population", fill="") +
          coord_flip()

# plot it
myplot

```
*Don't worry if you don't get all the ggplot code now. It's here for reference.*



## Data Wrangling

Your R skills can help you reformat the data and make it more usable.

Let's fetch population data for 2010 and 2000 by state.

Then we will combine these into one data frame using the `tidyverses::bind_rows` function

```{r}
# Fetch 2000 population data by state
pop2000 <- get_decennial(geography = "state",
                         variables = c(pop2000="P001001"), 
                         year = 2000)

# Fetch 2010 population data by state
pop2010 <- get_decennial(geography = "state",
                         variables = c(pop2010="P001001"), 
                         year = 2010)

# Use tidyverse `bind_rows` function to combine the data for these years
state_pop <- bind_rows(pop2000, pop2010)

# Take a look with head or View
head(state_pop)
```

# Any Questions?

# Part 2. Mapping


## Mapping Census Data with `tidycensus`

You can fetch geographic data by adding the parameter **geometry=TRUE** to `tidycensus` functions

- Under the hood, tidycensus calls the `tigris` package to fetch data from the Census Geographic Data APIs.

- Only a subset of data available via `tigris` can be accessed via `tidycensus`.

You can then use your favorite R mapping functions or libraries like `plot`, `ggplot`, and `tmap` to make maps.

## Geometry Options

Before fetching census geographic data, we need to set the option `tigris_use_cache` to TRUE

Caching greatly speeds things up if you fetch the same census geographic data repeatedly.

```{r}
# Tigris options - used by tidycensus
# Cache retrieved geographic data locally
options(tigris_use_cache = TRUE)  
```


## Fetch Geographic Boundary Data with `tidycensus`

We fetch the geospatial data by setting **geometry=TRUE**.

```{r}
pop2010geo <- get_decennial(geography = "state", 
                          variables = c(pop10="P001001"), 
                          year = 2010, 
                          output="wide", 
                          geometry=TRUE) # Fetch geometry with the data for mapping
 
```

## Take a look

Let's take a minute to discuss the format of an `sf` spatial object.
```{r}
head(pop2010geo, 3)
```


## Geospatial Data in R

R `sf` objects include

- a dataframe with a `geometry` column with the name  `geometry`

    - The geometry can be of type POINT, LINE, POLYGON
    - or, MULTIPOINT, MULTILINE or MULTIPOLGYON

- a `CRS` (coordinate reference system), specified by
    - epsg(SRID) code
    - proj4string
    
For a deeper understanding of the `sf` package and its functionality, we recommend 

- our [Geospatial-Fundamentals-in-R-with-sf](https://github.com/dlab-berkeley/Geospatial-Fundamentals-in-R-with-sf) workshop

- the [Geocomputation with R ebook](https://geocompr.robinlovelace.net/)
    
## Census Data Coordinate Reference System (CRS)

All census geographic data use the `NAD83` CRS, or coordinate reference system. `NAD83` stands for North American Datum of 1983. The geographic coordinates are longitude and latitude values encoded as decimal degrees.

`WGS84`, or [The World Geodetic System of 1984](https://en.wikipedia.org/wiki/World_Geodetic_System) is the most commonly used geographic CRS. The difference between point locations on the surface of the earth in these two systems can vary up to 1 meter in continental US.

Many geospatial operations require you transform data to a common CRS before conducting spatial analysis or mapping.  

- This could be an issue if you try to combine the census geospatial data with other geospatial data.

An in-depth discussion of CRSs is outside the scope of this workshop. See [Geocomputation in R](https://geocompr.robinlovelace.net/reproj-geo-data.html) for more information.

## Mapping `sf` Spatial Objects

We can use `plot` to make a quick map the geometry stored in an `sf` spatial object.

```{r}
plot(pop2010geo$geometry)
```

## Question

What do you get if you plot the `sf` object without specifying "$geometry"

Try it!
```{r, eval=F}
plot(pop2010geo)
```

## The Challenge of US maps

The vast geographic extent and non-contiguous nature of the USA makes it difficult to map.

```{r, echo=F}
plot(pop2010geo$geometry) #view again
```

## Fetch Geographic Data with `tidycensus`, SHIFTED

tidycensus includes a `shift_geo` parameter to shift AK & HI to below Texas.
```{r}

pop2010geo_shifted <- get_decennial(geography = "state", 
                                    variables = c(pop10="P001001"), 
                                    output="wide",
                                    year = 2010, 
                                    geometry=TRUE, 
                                    shift_geo=TRUE)

```

## Shift Happens!
```{r}
plot(pop2010geo_shifted$geometry)
```

## Save it

You can save any `sf` data object to a shapefile using `st_write`

```{r, eval=F}
st_write(pop2010geo_shifted, "data_out/usa_pop2010_shifted.shp")
```

## Check it out

```{r, eval=F}

# Check to see if the data was written out to a shapefile
dir("data_out") 
```

## Mapping Data Values

You can use the sf `plot` command to make a map that sets the color of the geometry by the data values 

```{r}
plot(pop2010geo_shifted['pop10'])  # a choropleth map!
```

## `ggplot2` Map

```{r}
ggplot(pop2010geo_shifted, aes(fill = pop10)) + 
  geom_sf()  # tells ggplot that geographic data are being plotted
```

## Let's make that map prettier
```{r}
ggplot(pop2010geo_shifted, aes(fill = pop10)) + 
  geom_sf(color=NA) + # What does color=NA do
  coord_sf(crs = 3857) + # Dynamically change the CRS
  scale_fill_viridis_c(option = "viridis")  # Change the color palette
                                            # Try different options, e.g.
                                            # plasma, magma, inferno, cividis
```

## Challenge 

Create a `map` of CA Median Age by county in 2010
```{r}
# add your code below - solution in Solutions.R
## Fetch data + geometry for CA Median Age by county in 2010

## Map the data with the plot function

## Map the data with ggplot, setting the CRS to 3310 (CA Albers)
## which is the preferred CRS for statewide maps of CA

```


## Fetch Census Data and Geometry for Multiple States or Counties

We can fetch Census data and the **geometry** for more than one state or county with same function call.

- This is so much easier than any alternative approach!

- It can be applied to any available geographic tabulation areas (eg states, counties, tracts, places).

Let's try it with Census Tracts!

Fetch tract population and geometry data for Alameda and Contra Costa Counties.

```{r}
bay_counties <- c("Alameda", "Contra Costa", "Marin", "San Francisco",
                  "Sonoma", "Napa","Solano", "San Mateo", "Santa Clara")

bayarea_pop10 <- get_decennial(geography = "tract", 
                      variables = "P001001", 
                      year = 2010, 
                      state='CA',
                      county=bay_counties,
                      geometry=T)

# take a look
head(bayarea_pop10)

# Quick map
plot(bayarea_pop10['value'])
```

# Questions?

# Part 3. ACS 5-year data

## ACS Data with `get_acs`

We can use the tidycensus function  `get_acs` to retrieve `ACS data` using code very similar to `get_decennial`.

**BUT** the workflow is more complex because:

1. The ACS has a lot more tables and variables, and

2. The ACS contains **sample data**, so each ACS variable that you retrieve with `tidycensus` will fetch both an **estimate** of the value and a **margin of error**.

## ACS Data Products

The ACS has two primary data products - the ACS 1 year database and the 5 year database. 

- The ACS 3 year data product has been discontinued.
 
- The ACS 1 year data is more current but has a larger margin of error and is not available for Census geographies with a population of < 65,000.
  
- So the ACS 5-year data is the most commonly used data set.

## Fetch List of ACS 5-Year Variables

Let's use the `load_variables` function to get a dataframe of all variables from the ACS 2016—2020 5-year dataset.

- Then, find the variable name for `median household income`.

> The 2016-2020 American Community Survey 5-year estimates are scheduled to be released on Thursday, December 9, 2021.

```{r}
vars_acs2019 <- load_variables(year=2019, # the end year of the 2016-2020 period
                              dataset = 'acs5', # the ACS data product
                              cache = T) # Save locally for future access

# how many variables?
dim(vars_acs2019)

# Take a look at the resultant dataframe
## What is the variable for median household income?
#View(vars_acs2019)
```

## Fetch Data on Median Household Income

Let's fetch the median household income data for Alameda County by Census Tract.

```{r}

alco_mhhincome <- get_acs(geography='tract',
                        variables=c(median_hhincome = "B19013_001"),
                        year = 2019,
                        state='CA',
                        county='Alameda',
                        geometry=TRUE
                        )
```

## Take a look
```{r}
head(alco_mhhincome)
```


## Map Median Household Income by tract

What is the variable?
```{r}
plot(alco_mhhincome['estimate'])
```

## Fetching Multiple ACS-5 Variables

First identify the variables of interest.
```{r}
# Median household income by race/ethnicity: Variables from ACS 2015—19
#All households =   "B19013_001",
inc_by_race <- c(White = "B19013H_001",
                 Black = "B19013B_001",
                 Asian = "B19013D_001",
                 Hispanic = "B19013I_001" )
```

## Fetch the Data

Fetch census tract data for multiple variables at once.
```{r}
alco_mhhinc_by_race <- get_acs(geography='tract',
                                  variables=inc_by_race,
                                  year = 2019,
                                  state='CA',
                                  county='Alameda',
                                  geometry=T )
```

##  Facet Map
Facet maps are a way to create visualizations of `small multiples`, or subsets of the data in order to facilitate comparisons. Here, we use ggplot's `facet_wrap` function to make multiple maps of median household income by race for Alameda County.

```{r}
# Create the map
medhhinc_facet_map <- alco_mhhinc_by_race %>%
                        ggplot(aes(fill = estimate)) +
                          facet_wrap(~variable) +
                          geom_sf(color=NA) +
                          scale_fill_viridis_c(option="magma")

# Display the map
medhhinc_facet_map
```

## Challenge

Make a ggplot map of MEDIAN GROSS RENT in San Francisc County by tract
using data from the ACS 2015—2019 5-year product.
```{r, eval=F}
# Fill in the code to fetch the data - (Solutions.R has the code)

# Median household rent for San Francisco County
alcc_medrent <- get_acs(geography= ,
                              variables= ,
                              year = ,
                              state= ,
                              county= ,
                              geometry=)

```



# Interactive Mapping

## Interactive Mapping

Interactive mapping gives the RStudio environment some of the functionality of desktop GIS.

There are a number of R packages tat you can use, including:

- [mapview](https://r-spatial.github.io/mapview/): quick interactive exploratory data viewing
  
- [tmap](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html): great static and interactive maps
  
- [Leaflet](https://rstudio.github.io/leaflet/): highly customizable interactive maps

*All of these are based on the Leaflet Javascript Library.*

## Mapview

Let's use mapview to make some quick interactive maps of our median hhousehold income data

```{r, eval=F}
mapview(alco_mhhincome)
```

## Interactive Choropleth Map
```{r, eval=F}
mapview(alco_mhhincome, zcol="estimate")
```


## Challenge

Use `mapview` to create a map of median household income (`alcc_medrent`).

## Challenge Solution
```{r, eval=F}
mapview(alcc_medrent, zcol='estimate')
```

# Any Questions?

## Figuring out the ACS Variables to use

ACS variables can be confusing. 

Some ways to identify the best variables to explore:

- Web search, especially Census web resources, can help.

- The Census Reporter website (https://censusreporter.org) provides another tool for navigating topics, tables, and variable names.

- The NHGIS website (nhgis.org) is a great way to browse variables of interest.

 
## Margins of Error (MOE)

We haven't talked about it but it may be important in your work with ACS data.

Math is needed to combine MOEs when you combine variables.

- `tidycensus` includes some nice [functions](https://walker-data.com/tidycensus/articles/margins-of-error.html) for these calculations and a good overview of the topic.

## Summary

`tidycensus` offers two key functions for fetching census tabular and geographic: **get_acs** and **get_decennial**.  

- The `load_variables` function helps identify the names of census variables of interest.

Support for fetching `population estimates` and `migration flow` census data was recently added to tidycensus. You can read up on it on the [tidycensus documentation website](https://walker-data.com/tidycensus/articles/other-datasets.html)

Using `tidycensus` to fetch the tabular data or both tabular and geographic data is IMO way easier than any alternatives, **IF** you (1) know R, (2) know a bit about working with geographic data in R.

This approach is also scaleable if you want multiple census variables for various locations and tabulation areas.

You can greatly enhance your maps if you make them with `ggplot2` rather than the default `plot` command.

Interactive mapping greatly enhances your ability to do exploratory data analysis in RStudio.


# References

Much of this tutorial is based on resources by Kyle Walker, author of `tidycensus`. See:

- [tidycensus webpage](https://walker-data.com/tidycensus/index.html))
- [Analyzing the US Census with R](https://walker-data.com/census-r), an online book.

Related D-Lab Workshops

- [R Fundamentals](https://github.com/dlab-berkeley/R-Fundamentals)
- [Geospatial Data in R, parts 1, 2, & 3](https://github.com/dlab-berkeley/R-Geospatial-Fundamentals)

Great online resource for working with spatial data in R

- [Geocomputation with R](https://geocompr.robinlovelace.net/)


# Appendix

## A More Complex Query

Let's use the 2010 census data to map the percent of San Francisco (SF) properties that were rented.

To start, identify the variables for the

- Total number of housing units 

- Number of renter occupied units

## Complete the query
```{r, eval=F}
sf_rented <- get_decennial(geography =  ,  # census tabulation unit
                           variables =   , # number of households rented
                           year =  , 
                           summary_var = ,  # Total households
                           state=,
                           county=,
                           geometry=)
```

##  and here it is SF Percent Rented Units, 2010 

```{r}
sf_rented <- get_decennial(geography = "tract",  # census tabulation unit
                           variables =  "H004004", #number of households rented
                           year = 2010, 
                           summary_var = "H004001",  # Total households
                           state='CA',
                           county='San Francisco',
                           geometry=T)

# take a look at the output
head(sf_rented)
```


## Calculate Percent Rented

```{r}
sf_pct_rented <- sf_rented[sf_rented$value > 0,] %>%
                 mutate(pct = 100 * (value / summary_value))

# Take a look
head(sf_pct_rented)

```

## Map the result
```{r}
plot(sf_pct_rented['pct'])
```
