---
title: "Census Data Wrangling and Mapping in R"
author: "Patty Frontiera"
date: "04/01/2022"
theme: readable
output: 
  html_document: 
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Census Data and `tidycensus`

In this tutorial, we will work through several exercises using the [tidycensus](https://walker-data.com/tidycensus/index.html) R package to fetch, wrangle, and map census data.

The key `tidycensus` functions we will use today are:

-   `census_api_key`: makes your Census API key available to tidycensus

-   `load_variables`: retrieves a dataframe of available census data variables

-   `get_decennial`: fetch census data from a recent decennial censuses - 2000, 2010 (and soon 2020)

-   `get_acs`: fetch census data from an ACS (American Community Survey) 1 or 5 year dataset, 2005 - 2020.

# Setup

Be sure to **clone** or **downloaded and unzip** the workshop files from: <https://github.com/dlab-berkeley/Census-Data-in-R>

Then:

1.  Open the folder with the workshop files

2.  Double-click on the R Project file `Census-Data-in-R.Rproj`

3.  This should open RStudio - with the `Files` panel displaying the workshop folder contents.

4.  Double-click on the file `Census-Data-in-R.Rmd` to follow along!

*You can also click on the file `Census-Data-in-R.html` in the Files tab to open the workshop tutorial in a web brower.*

## Install packages

> If you installed any of these packages awhile ago, (especially `tidycensus`), it's a good idea to install updates when you can (though not during the workshop as things can break!).

```{r}

# Uncomment this to install packages, if necessary.
# install.packages(c("here", "tidyverse", "sf", "leaflet", "mapview", "tigris", "tidycensus"))

library(here)
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(tigris)
library(tidycensus)

```

These seven libraries should be loaded in your environment now.

```{r}

# If you run this chunk, output from the "here" function should be visible below. This is your local directory path. We can use this to import files later on. 
here()

```

## Census API Key

You need a `Census API key` to programmatically fetch census data.

-   Get it here (pretty quickly): <https://api.census.gov/data/key_signup.html>

-   The key will be sent to your email and you will need to click to activate it.

-   Keep the email with the key open for use in this workshop.

*For more info on all available Census APIs see: <https://www.census.gov/data/developers/data-sets.html>*

### Add Your Census API Key

To use your Census API Key in R

1.  Copy and paste your Census API key from your email

2.  Use the tidycensus function **census_api_key** to register your API key with tidycensus. **Don't forget to put quotes around the key!**.

```{r, eval=FALSE}

# Install your census api key - long alphanumeric string
census_api_key("THE_BIG_LONG_ALPHANUMERIC_API_KEY_YOU_GOT_FROM_CENSUS")

```

Another way to add your Census API Key:

I keep my key in a file so no one can see it. One way to do this is by making a script that creates a variable key, and then using the `source` function to add that script as an object into your coding environment. The code chunk below is an example of how you might do that:

```{r,eval=FALSE}

# source (run) an r script that creates a variable with my key
#source("/Users/pattyf/Documents/Dlab/workshops/keys/census_api_key.R")

#print(my_census_api_key) 

# register the key
census_api_key(key = my_census_api_key)
```

# Decennial Census Data

## The `get_decennial` function

We start by fetching `total population` from the 2010 Census with tidycensus's `get_decennial` function. Let's first talk about the code.

```{r}
pop2010 <- get_decennial(geography = "state",   # census tabulation unit
                         variables = "P001001", # variable(s) of interest
                         year = 2010)           # census year

head(pop2010)          
```

## Fetching data for more than one Census variable

We can pass a vector of census identifiers to the `variables` function argument if we want to get data for more than one variable. Below we add `P0002002` for population in urban areas.

```{r}
pop2010 <- get_decennial(geography="state",
                         variables = c("P001001","P002002"), # variable(s) of interest
                         year = 2010)           # census year

# take a look
head(pop2010)
```

We can see the data for both variables if we sort the output by county name.

```{r}
# Sort dataframe by county names (the NAME column)
pop2010 %>% arrange(NAME) %>% head()
```

## tidycensus returns `tidy` data

By default, tidycensus returns data in a `tidy`, or `long` format that allows data for multiple variables to be contained within the `variable` and `value` columns.

This is in contrast to untidy, or `wide` data where each variable is in its own column.

tidycensus can return `wide` data if you can add the parameter `output=wide` to the function call.

```{r}

# wide format
pop2010w <- get_decennial(geography = "state",   # census tabulation unit
                         variables = c("P001001","P002002"), # variable(s) of interest
                         year = 2010,           # census year
                         output="wide")         # get output in wide format
head(pop2010w) 
```

### The `GEOID` column

The `GEOID` column is included in tidycensus output by default.

This is a Census geographic identifier for the tabulation unit.

The `GEOID` is sometimes called the Census `FIPS` code and for most tabulation units these are the same.

The `GEOID` makes it possible to link to Census demographic data to Census geographic data and make maps. We will do this in a bit.

The `GEOID` is a text string and must be quoted.

-   Beware of GEOID leading zeros, since some software will remove these and convert GEIOD values to numbers (rather than text strings).

> Question: **What is the GEOID for California?**

# Census Tabulation Units

Public census data is typically aggregated by census geographies to protect privacy.

These census geographies are called `Census tabulation units`.

-   Some of these are real administrative units like states and counties.
-   Others are statistical units created by the census, like census tracts and block groups.

Some of the most common geographic tabulation units and their tidycensus function abbreviations are shown below, along with **required** and available filters that limit what data are returned.

| Geography     | Definition           | Filter(s)             | Used in tidycensus         |
|-----------------|-----------------|-----------------|---------------------|
| "us"          | United States        |                       | get_acs(), get_decennial() |
| "region"      | Census region        |                       | get_acs(), get_decennial() |
| "state"       | State or equivalent  | state                 | get_acs(), get_decennial() |
| "county"      | County or equivalent | state, county         | get_acs(), get_decennial() |
| "place"       | Census place         | state                 | get_acs(), get_decennial() |
| "tract"       | Census tract         | **state**, county     | get_acs(), get_decennial() |
| "block group" | Census block group   | **state**, county     | get_acs(), get_decennial() |
| "block"       | Census block         | **state**, **county** | get_decennial() only!      |

## `get_decennial` Tabulation Units and Filters

Let's take a few minutes to practice fetching population data with the `get_decennial` function.

-   See `?get_decennial` for help

### Challenge 1

> Open **Census-Data-in-R-Challenges.Rmd** and use the `get_decennial` function like we've seen above to fetch population data. *Solutions are available in the Solutions folder, as needed.*

### Changing the tabulation unit

Let's fetch 2010 population data for CA counties

**What changes in the code?**

```{r}

get_decennial(geography = "county",              # census tabulation unit
                          variables = "P001001", # variable(s) of interest
                          year = 2010,           # census year
                          state='CA')            # Filter by state is CA
```

#### Questions

-   How do we specify the state of CA above? How else can we?
-   Can you fetch population data for all counties in the USA or do you need to have a `state=` filter?

### Adding a county filter

You can also filter tidycensus results by `county`

```{r}

get_decennial(geography = "county",              # census tabulation unit
                          variables = "P001001", # variable(s) of interest
                          year = 2010,           # census year
                          state='CA',            # Filter by state is CA
                          county='Alameda')      # Filter by county Alameda
```

### Challenge 2

> In **Census-Data-in-R-Challenges.Rmd**, alter the code above to fetch 2010 population for Alameda & San Francisco Counties. Then try Challenge 2B & 2C.

## Visualizing Results

We can visualize data to get a quick overview of the distribution of the values.

It's a first step in exploratory data analysis and a last step in data communication.

`ggplot2` is the most commonly used R package for data visualization.

-   It is loaded when you load the `tidyverse` package.

Let's use it to visualize the population data.

### Plot 2010 Population by state

Use `ggplot2` to create an ordered horizontal bar chart.

```{r}

# create a plot.
pop_plot <- ggplot(data=pop2010, 
                   # set aesthetic variables
                   aes(x=value/1000000, y=reorder(NAME,value)) ) + 
                   # pick geometry
                   geom_bar(stat="identity") +  
                   # add theme and titles. 
                   theme_minimal() + 
                   labs(title = "2010 US Population by State") +
                   xlab("Population (in Millions)") +
                   ylab("State")

# display the plot.
pop_plot

```

Developing your `ggplot2` knowledge can really enhance your data analysis skills.

In combination with `tidycensus` it creates a powerful, reproducible data science workflow.

# Identifying Census Variables

In the code above we fetched data for total population in 2010 using the variable `"P001001"`.

That is not an obvious variable name, so how do we get those census data identifiers?

We can use the tidycensus `load_variables` function for this.

## `load_variables` function

For any census dataset like the decennial census or the ACS 1 or 5-year estimates, use the `load_variables` function to fetch all available variables and identifiers.

Since these datasets have many, many variables, save the resulting dataframe to a variable and cache it locally so you do not need to repeatedly retrieve it over the web.

```{r}

vars2010 <- load_variables(year=2010,        # Year or end year for ACS-5yr
                           dataset = 'sf1',  # 'sf1' for decennial census
                           cache = TRUE)     # Save fetched data locally

# How large is the output
dim(vars2010)

# Take a look with head or View
head(vars2010)

```

## 2010 Decennial Census Tables & Variables

-   Over 3,000 unique variables that describe population and housing characteristics

-   Organized in `333` Tables

    -   177 population tables (identified with a ''P'') available to the block level
    -   58 housing tables (identified with an ''H'') available to the block level
    -   82 population tables (identified with a ''PCT'') available to the census tract level
    -   4 housing tables (identified with an "HCT") available to the census tract level
    -   10 population tables (identified with a "PCO") available to the county level
    -   plus 2 additional PCT tables

<small>See: <https://www.census.gov/data/datasets/2010/dec/summary-file-1.html></small>

#### What Variable Has the 2010 Total Population value?

We know this from our previous code blocks, but let's find it for practice navigating the dataframe.

-   Let's sort and filter the `vars2010` dataframe to find it.

#### *Questions:*

What 2010 decennial census variable contains...

-   `Median Age`

-   `Average Family Size`

-   `Number of occupied housing units`

\*See **Census-Data-in-R-Solutions.Rmd** if needed (under Variable Questions)

### Challenge 3

> Return to **Census-Data-in-R-Challenges.Rmd** and use the `get_decennial` function to fetch and plot an `Avg Family Size`varaible by CA County in `2010`, and name the call as a dataframe, `ca_fam_size`. Once you've done that, plot the dataframe with the `ggplot`call below. **Hint: "P037001"**

### Challenge 4

> Repeat the previous challenge with data from the `2000` decennial census. Don't assume variable names are the same across the 2000 and 2010 census

> **Use `load_variables` to check the variable name!**

# Census Tract Data

Census tracts are the most commonly used census tabulation unit.

Let's fetch population data for the census tabulation unit to *tract*

> Because of the large number of census tracts, you **MUST** specify a state when requesting these data with tidycensus.

```{r}
## Fetch population by **tract** for California.
ca_tract_pop2010 <- get_decennial(geography = "tract",   # census tab unit
                                   variables = "P001001", # var of interest
                                   year = 2010,           # census year
                                   state='CA')      # State filter

# How many tracts in CA
dim(ca_tract_pop2010)

# take a look
head(ca_tract_pop2010)
```

## Fetching Census Tract Data

Census tract data can be quite large!

Fortunately, you can also limit the results to one or more counties.

```{r,}

tract_pop2010 <- get_decennial(geography = "tract",   # census tabulation unit
                         variables = "P001001",       # variable of interest
                         year = 2010,                 # census year - only one!
                         state="CA",                  # limit to California
                         county=c("Alameda","Contra Costa"))  # & counties

dim(tract_pop2010)

```

## Customizing tidycensus output

What **two** things are new here?

```{r}

#urban and rural pop for 3 CA counties
ur_pop10 <- get_decennial(geography = "county",  # census tabulation unit
                           variables = c(urban="P002002",rural="P002005"),
                           year = 2010, 
                           summary_var = "P002001",  # The denominator
                           state='CA',
                           county=c("Napa","Sonoma","Mendocino"))

```

#### When fetching census data...

We have already specified more than one variable:

    variables = c("P002002","P002005")

1.  You can also rename the values in the output 'variable' column.

<!-- -->

    variables = c(urban="P002002",rural="P002005")

2.  You can identify a `summary_var` (a denominator - here, the total count of all people or households surveyed. Can be used for calculations like percent of total.)

<!-- -->

    summary_var = "P002001"

#### Now let's take a look at the resultant dataframe

```{r}
# take a look at the results
ur_pop10
```

#### Calculating Percents

The `summary_value` column comes in handy when you want to compute percent of total, for example:

```{r}

# Calculate the percent of population that is Urban or Rural
ur_pop10 <- ur_pop10 %>%
            mutate(pct = 100 * (value / summary_value))

# Take a look at the output.
ur_pop10 

```

A plot gives us compact visual summaries of the data.

```{r}

## Plot it with ggplot2
myplot <- ggplot(data = ur_pop10, 
          mapping = aes(x = NAME, fill = variable, 
                     y = ifelse(test = variable == "urban", 
                                yes = -pct, no = pct))) +
          geom_bar(stat = "identity") +
          scale_y_continuous(labels = abs, limits=c(-100,100)) +
          labs(title="Urban & Rural Population in Wine Country", 
               x="County", y = " Percent of Population", fill="") +
          coord_flip()

myplot

```

Don't worry if you don't get all the ggplot code now. It's here for reference.

-   You may want to check out D-Lab's *R Data Visualization with ggplot* workshop!

## Data Wrangling to Combine Data from 2 Censuses

You can use your R skills to reformat the data and make it more usable.

Let's fetch population data for 2010 and 2000 by state.

Then we will combine these into one data frame using the `tidyverses::bind_rows` function

```{r}

# Fetch 2000 population data by state
pop2000 <- get_decennial(geography = "state",
                         variables = c(pop2000="P001001"), 
                         year = 2000)

# Fetch 2010 population data by state
pop2010 <- get_decennial(geography = "state",
                         variables = c(pop2010="P001001"), 
                         year = 2010)

# Use tidyverse `bind_rows` function to combine the data for these years
state_pop <- bind_rows(pop2000, pop2010)

# Take a look with head or View
state_pop %>% arrange(NAME) %>% head(10)
```

# Saving `tidycensus` output

The data we fetch using `tidycensus` is stored in an R dataframe.

We can use the `write.csv` or `write_csv` function to save the contents of a dataframe to a CSV file.

```{r}
write_csv(state_pop, here('data_out/state_pop_2010.csv') )
```

> **Any Questions?**

# Mapping Census Data

You can fetch census geographic data by adding the parameter **geometry=TRUE** to `tidycensus` functions

-   Under the hood, tidycensus calls the `tigris` package to fetch data from the Census Geographic Data APIs.

You can then use your favorite R mapping packages like `sf`, `ggplot`, `tmap`, `mapview` and `leaflet` to make maps.

## Geometry Options

Before fetching census geographic data, we need to set the option `tigris_use_cache` to TRUE

Caching saves data locally. This greatly speeds things up if you fetch the same census geographic data repeatedly.

```{r}

# Tigris options - used by tidycensus
# Cache retrieved geographic data locally
options(tigris_use_cache = TRUE)  

```

## Fetch Geographic Boundary Data with `tidycensus`

We fetch the census geographic data by setting **geometry=TRUE**.

```{r}

pop2010geo <- get_decennial(geography = "state", 
                          variables = c(pop10="P001001"), 
                          year = 2010, 
                          output="wide", 
                          geometry=TRUE) # Fetch geometry data for mapping
 
```

### Take a look

Let's take a minute to discuss the format of an `sf` spatial object.

```{r}

head(pop2010geo, 3)

```

# Geospatial Data in R

The `tidycensus` package uses the R `sf` package to manage geospatial data.

R `sf` objects include:

-   a dataframe with a `geometry` column labeled `geometry`

    -   The geometry can be of type POINT, LINE, POLYGON
    -   or, MULTIPOINT, MULTILINE or MULTIPOLGYON

-   a `CRS` (coordinate reference system), specified by

    -   epsg(SRID) code
    -   proj4string

For a deeper understanding of the `sf` package and its functionality, we recommend

-   our [R-Geospatial-Fundamentals](https://github.com/dlab-berkeley/R-Geospatial-Fundamentals) workshop

-   the [Geocomputation with R ebook](https://geocompr.robinlovelace.net/)

## Census Data Coordinate Reference System (CRS)

All geospatial data are referenced to the surface of the earth with a `CRS`, or `coordinate reference system`. Anyone working with geospatial data will need to develop an understanding of CRSs.

Fortunately, many of us are familiar with longitude and latitude, which are geographic coordinates. But there are different versions of geographic CRSs. And there are also projected CRSs which transform longitude and latitude to 2 dimensional surface for mapping & analysis.

All census geographic data use the `NAD83` geographic CRS. `NAD83` stands for North American Datum of 1983. This CRS (or version of latitude and longitude) is best for locations in North America.

Many geospatial operations require you transform data to a common CRS before conducting spatial analysis or mapping.

-   This could be an issue if you try to combine the census geospatial data with other geospatial data. But it is not an issue in this tutorial.

An in-depth discussion of CRSs is outside the scope of this workshop. See [Geocomputation in R](https://geocompr.robinlovelace.net/reproj-geo-data.html) for more information.

## Mapping `sf` Spatial Objects

We can use `sf::plot` to make a quick map the geometry stored in an `sf` spatial object.

```{r}
# plot the geometry column data
plot(pop2010geo$geometry)
```

### The Challenge of US maps

The vast geographic extent and non-contiguous nature of the USA makes it difficult to map.

Fortunately, tidycensus includes a `shift_geo` parameter to shift AK & HI to below Texas.

```{r}

pop2010geo_shifted <- get_decennial(geography = "state", 
                                    variables = c(pop10="P001001"), 
                                    output="wide",
                                    year = 2010, 
                                    geometry=TRUE, 
                                    shift_geo=TRUE)

## Shift Happens!
plot(pop2010geo_shifted$geometry)
```

## Saving Spatial Objects

You can save any `sf` spatial data object to an [ESRI shapefile](https://en.wikipedia.org/wiki/Shapefile) using `st_write`

```{r, eval=F}

st_write(pop2010geo_shifted, here("data_out/usa_pop2010_shifted.shp"))

```

Now take a look at the output shapefile.

```{r, eval=F}

# Check to see if the data was written out to a shapefile
dir(here("data_out")) 
```

### ESRI Shapefiles

You can see from this output that an ESRI shapefile is actually a collection of files that all have the same prefix.

Shapefiles are the most common file format for geospatial data. So it's worthwhile to learn more about them if you will be working with census geographic data.

-   Check out the [Shapefile Wikipedia page](https://en.wikipedia.org/wiki/Shapefile) for more information.

## Mapping Data Values

You can use the sf `plot` command to make a map that sets the color of the geometry by the data values

-   This type of map is called a `thematic map`.

-   When the features being plotted are areas (or polygons), it's called a `choropleth` map!

```{r}
# Name the column with the variable values to make
# a thematic map, also called a choropleth map.
plot(pop2010geo_shifted['pop10'])  

```

### `ggplot2` Map

`ggplot` knows how to map sf objects!

```{r}

ggplot(pop2010geo_shifted, aes(fill = pop10)) + 
  geom_sf()  # tells ggplot that geographic data are being plotted

```

If you are familiar with `sf` objects and `ggplot` you can further customize your maps.

```{r}

ggplot(pop2010geo_shifted, aes(fill = pop10)) + 
  geom_sf(color=NA) + # What does color=NA do
  coord_sf(crs = 3857) + # Dynamically change the CRS
  scale_fill_viridis_c(option = "viridis")  # Change the color palette
                                            # Try different options, e.g.
                                            # plasma, magma, inferno, cividis

```

### Challenge 5

> In your **Census-Data-in-R-Challenges.Rmd** file, create a map of `Median Age by California County in 2010`. *Solutions are in the Census-Data-in-R-Solutions.Rmd file*

### Fetch Census Data and Geometry for Multiple States or Counties

We can fetch Census data and **geometry** for more than one state or county with same function call.

-   This is so much easier than any alternative approach!

-   It can be applied to any available geographic tabulation areas (eg states, counties, tracts, places).

Let's try it with Census Tracts!

Fetch tract population and geometry data for Bay Area Counties.

```{r}

bay_counties <- c("Alameda", "Contra Costa", "Marin", "San Francisco",
                  "Sonoma", "Napa","Solano", "San Mateo", "Santa Clara")

bayarea_pop10 <- get_decennial(geography = "tract", 
                      variables = "P001001", 
                      year = 2010, 
                      state='CA',
                      county=bay_counties,
                      geometry=T)

# Quick map
plot(bayarea_pop10['value'])
```

> **Any Questions?**

# Fetching ACS data with `get_acs`

-   ACS data contains the most recent information about the American population.

-   We can use the tidycensus function `get_acs` to retrieve `ACS data` using code very similar to `get_decennial`.

**BUT** the workflow is more complex because:

1.  ACS data has a lot more variables, and

2.  ACS data are **sample data**, so each ACS variable that you retrieve with `tidycensus` will fetch both an **estimate** of the value and a **margin of error**.

## ACS Data Products

The ACS has two primary data products - the ACS 1 year estimates and the 5 year estimates.

-   The `ACS 1 year estimates` are more current but hasve a larger margin of error and is not available for Census geographies with a population of \< 65,000.

-   The `ACS 5-year estimates` are more stable but represent a larger time period.

-   *The `ACS 3 year estimates` has been discontinued.*

## Fetch metadata on ACS 5-Year Variables

Let's use the `load_variables` function to get a dataframe of all variables from the ACS 2016---2020 5-year dataset.

-   Note: we change the dataset value to `acs5` where before we used `sf1` to fetch info on the decennial census variables.

```{r}

vars_acs2020 <- load_variables(year=2020,       # end year 2016-2020 period
                              dataset = 'acs5', # the ACS data product
                              cache = T)        # Save locally for future use

# how many variables?
dim(vars_acs2020)
```

### Exploring the ACS Variables

View the `vars_acs2020` dataframe to find the variable name for `median household income`.

```{r, eval=FALSE}
#View(vars_acs2020)
```

> *Question* Is the variable name for `total population` in the ACS 5 year 2020 data the same as it is in the 2010 census data?

### Fetch ACS Data on Median Household Income

Let's fetch the median household income data for San Francisco County by census tract.

```{r}

med_hhincome <- get_acs(geography='tract',
                        variables="B19013_001",
                        year = 2020,
                        state='CA',
                        county='San Francisco',
                        geometry=TRUE # get the geography too
                        )
```

Take a look at the output

```{r}
head(med_hhincome)
```

### ACS Output

The census data returned by the `get_acs` function is a bit different from that returned by `get_decennial`.

-   What is the name of the variable containing the income data?

-   What is the name if we set `output="wide"`?

```{r}
med_hhincome_wide <- get_acs(geography='tract',
                        variables="B19013_001",
                        year = 2020,
                        state='CA',
                        county='San Francisco',
                        geometry=TRUE, # get the geography too
                        output="wide"
                        )
                        
# uncomment and run to view
# head(med_hhincome_wide)
```

### Map Median Household Income by tract

Use `sf::plot` to create a map of median household income in San Francisco.

```{r}
plot(med_hhincome['estimate'])
```

*What do you think of that map?*

It's odd because San Francisco County is not the same as the city of San Francisco and what we want to map is the city.

#### Create a map with ggplot

We can use ggplot to zoom in on the city by setting the x axis limits to a narrower geographic range.

```{r}
ggplot(med_hhincome, aes(fill = estimate)) + 
  geom_sf() +
  xlim(-122.55, -122.3)
  
```

### Question

> Why do you think we have NA values in the ACS estimates?

### Fetching Multiple ACS-5 Variables

We can drill down into the ACS data by fetching data for subgroups, where available.

Let's fetch median household income by race.

First identify the variables of interest.

```{r}

# Median household income by race/ethnicity: Variables from ACS 2015—19
# All households =   "B19013_001",
inc_by_race <- c(White = "B19013H_001",
                 Black = "B19013B_001",
                 Asian = "B19013D_001",
                 Hispanic = "B19013I_001" )
```

Fetch census tract data for multiple variables at once.

```{r}
# Fetch the Data
alco_mhhinc_by_race <- get_acs(geography='tract',
                                  variables=inc_by_race,
                                  year = 2019,
                                  state='CA',
                                  county='Alameda',
                                  geometry=T )
```

## Facet Mapping

Facet maps are a way to create visualizations of `small multiples`, or subsets of the data in order to facilitate comparisons. Here, we use ggplot's `facet_wrap` function to make multiple maps of median household income by race for Alameda County.

```{r}
# Create the map
medhhinc_facet_map <- alco_mhhinc_by_race %>%
                        ggplot(aes(fill = estimate)) +
                          facet_wrap(~variable) +
                          geom_sf(color=NA) +   # why color=NA?
                          scale_fill_viridis_c(option="magma")

# Display the map
medhhinc_facet_map
```

### Challenge 6

> In **Census-Data-in-R-Challenges.Rmd** file, Make a ggplot map of MEDIAN GROSS `RENT` in San Francisco County by tract using data from the ACS 2016---2020 5-year product. *Check Census-Data-in-R-Solutions.Rmd for answers, as needed.*

> **Any Questions?**

# Interactive Mapping

Interactive mapping gives the RStudio environment some of the functionality of desktop GIS.

There are a number of R packages that you can use, including:

-   [mapview](https://r-spatial.github.io/mapview/): quick interactive exploratory data viewing

-   [tmap](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html): great static and interactive maps

-   [Leaflet](https://rstudio.github.io/leaflet/): highly customizable interactive maps

*All of these are based on the `Leaflet Javascript Library`.*

### [mapview](https://r-spatial.github.io/mapview/)

Let's use `mapview` to make quick interactive maps of the median hhousehold income data

```{r}

mapview(med_hhincome)

```

When passed the name of an `sf` object and no other options, `mapview` will:

-   display the geometry using a single color for the fill and for the stroke

-   display the feature ID `on hover`

-   display the data from the dataframe `on click`

#### Mapview Thematic Maps

The `zcol` argument will take a column name and color the features by the values in that column.

```{r}

mapview(med_hhincome, zcol="estimate")

```

### Challenge 7

> In the **Census-Data-in-R-Challenges.Rmd**, use `mapview` to create an interactive map of median household rent.

## Determining what ACS Variables to use

ACS variables can be confusing.

Some ways to identify the best variables to explore:

-   Web search, especially Census web resources, can help.

-   The Census Reporter website (<https://censusreporter.org>) provides another tool for navigating topics, tables, and variable names.

-   The NHGIS website (nhgis.org) is a great way to browse variables of interest.

## ACS Margins of Error (MOE)

We haven't talked about it but it may be important in your work with ACS data.

Math is needed to combine MOEs when you combine variables.

-   `tidycensus` includes some nice [functions](https://walker-data.com/tidycensus/articles/margins-of-error.html) for these calculations and a good overview of the topic.

# Summary

`tidycensus` offers two key functions for fetching census tabular and geographic: **get_acs** and **get_decennial**.

-   The **load_variables** function helps identify the names of census variables of interest.

Support for fetching `population estimates` and `migration flow` census data was recently added to tidycensus. You can read up on it on the [tidycensus documentation website](https://walker-data.com/tidycensus/articles/other-datasets.html)

Using `tidycensus` to fetch the tabular data or both tabular and geographic data is IMO way easier than any alternatives, **IF** you (1) know R, (2) know a bit about working with geographic data in R.

This approach is also scaleable if you want multiple census variables for various locations and tabulation areas.

You can make publication or report ready maps with highly customizable `ggplot2` code or use the `sf::plot` command to make quick maps.

Interactive mapping greatly enhances your ability to do exploratory data analysis in RStudio.

### References

Much of this tutorial is based on resources by Kyle Walker, author of `tidycensus`. See:

-   [tidycensus webpage](https://walker-data.com/tidycensus/index.html))
-   [Analyzing the US Census with R](https://walker-data.com/census-r), an online book.

Related D-Lab Workshops

-   [R Fundamentals](https://github.com/dlab-berkeley/R-Fundamentals)
-   [Geospatial Data in R, parts 1, 2, & 3](https://github.com/dlab-berkeley/R-Geospatial-Fundamentals)

Great online resource for working with spatial data in R

-   [Geocomputation with R](https://geocompr.robinlovelace.net/)

------------------------------------------------------------------------
