---
title: "R-Census-Data-Maps-Challenges.Rmd"
author: "Avery Richards"
date: "3/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# install pacman for package management.
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  here, # locate files 
  tidyverse, # data wrangling
  sf,  # geospatial data management
  leaflet, # interactive mapping
  tigris, tidycensus  # census data 
)

```

These six libraries should be loaded in your environment now. 

```{r}

# If you run this chunk, output from the "here" function should be visible below. This is your local directory path. We can use this to import files later on. 
here()

```

*Solutions are available in the Solutions folder, as needed.* 

### Challenge 1

> Use the `get_decennial` function like we've seen above, but fill in the code arguments to fetch total population in 2010 just for California. 

```{r, challenge_1_code_A }

# add your code here. 

        
```

Alter the above code to fetch total pop in 2010 for CA, TX and FL
```{r, challenge_1_code_B }

# add your code here. 

```


### Challenge 2

> You can also filter tidycensus results by `county`. Alter the code below to fetch 2010 population for Alameda & San Francisco Counties.


```{r}

get_decennial(geography = "county",              # census tabulation unit
                          variables = "P001001", # variable(s) of interest
                          year = 2010,           # census year
                          state='CA',            # Filter by state is CA
                          county='Alameda')      # Filter by county Alameda
```



```{r}

# add your code here. 

```

What was the total population in the US in 2010?
```{r, challenge_2_code_B }

# add your code here. 

```

What census region had the largest population in the US in 2010?
```{r, challenge_2_code_C }

# add your code here. 

```

## Challenge 3

> Use the `get_decennial` function to fetch and plot an `Avg Family Size`varaible by CA County in `2010`, and name the call as a dataframe, `ca_fam_size`. Once you've done that, plot the dataframe with the `ggplot`call below. __Hint: "P037001"__ 

```{r, challenge_3A}

# add you code here to create a ca_fam_size dataframe with get_decennial.

```



```{r, challenge_3B}

# uncomment and plot from the ca_fam_size dataframe.
# ca_fam_size %>%
#   ggplot(aes(x = value, 
#              y = reorder(NAME, value))) + 
#       geom_point()

```

## Challenge 4

> Repeat the previous challenge with data from the `2000` decennial census. Don't assume variable names are the same across the 2000 and 2010 census

- Use `load_variables` to check!


```{r, challenge_4}
# Add your code below


```


## Challenge 5

Create a `map` of CA Median Age by county in 2010. 

```{r, challenge_5A}

# Add your code to fetch the data for the map. 

```


```{r, challenge_5B}

# Map the data with the plot function


```


```{r, challenge_5C}

# Map the data with ggplot, setting the CRS to 3310 (CA Albers), the preferred default CRS for statewide maps of CA.

```


## Challenge 6

> Make a ggplot map of MEDIAN GROSS RENT (`"B25064_001"`) in San Francisco County by tract
using data from the ACS 2016—2020 5-year dataset 


```{r, challenge_6}

# Add your code here.

```

## Challenge 7

>Use `mapview` to create an interactive map of SF median household rent.

```{r, challenge_7}

# Use mapview to create an interactive map of median household rent (from challenge 6).
# Your code here

```

> *Check Census-Data-in-R-Solutions.Rmd for answers, as needed.* 
