---
title: "snippets_to_save_for_later"
output: html_document
date: '2022-04-01'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### [Leaflet](https://rstudio.github.io/leaflet/):

Leaflet is the `ggplot2` of interactive mapping. Leaflet in R follows a tidyverse convention, using pipes (%>%) to create layers in the mapping object. We can use leaflet to create interactive maps allowing for more flexibility in design and features we can create. With added complexity in the code, of course! 
  
  ```{r}

# Create a color palette
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = med_hhincome$estimate
)

# specify dataset
leaflet(med_hhincome) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  # adjust color palette and ploygon features.
  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = .5,
              color = ~pal(estimate)) %>% 
  # add legend 
  addLegend(pal = pal, values = ~estimate,
            title = "Median Household Income",
            labFormat = labelFormat(prefix = "$"),
            position = "bottomleft")


```

---
  ## Appendix
  
  ### A More Complex Query
  
  Let's use the 2010 census data to map the percent of San Francisco (SF) properties that were rented.

To start, identify the variables for the

- Total number of housing units 

- Number of renter occupied units

### Complete the query
```{r, eval=F}

sf_rented <- get_decennial(geography =  ,  # census tabulation unit
                           variables =   , # number of households rented
                           year =  , 
                           summary_var = ,  # Total households
                           state=,
                           county=,
                           geometry=)
```

> And here it is SF Percent Rented Units, 2010 

```{r, eval=F}
sf_rented <- get_decennial(geography = "tract",  # census tabulation unit
                           variables =  "H004004", #number of households rented
                           year = 2010, 
                           summary_var = "H004001",  # Total households
                           state='CA',
                           county='San Francisco',
                           geometry=T)

# take a look at the output
head(sf_rented)
```


### Calculate Percent Rented

```{r, eval=F}
sf_pct_rented <- sf_rented[sf_rented$value > 0,] %>%
                 mutate(pct = 100 * (value / summary_value))

# Take a look
head(sf_pct_rented)

```

### Map the result
```{r, eval=F}
plot(sf_pct_rented['pct'])
```

